---
title: "ReproducibleResearchProject1"
author: "ChrisH"
date: "Sunday, January 18, 2015"
output: html_document
---

### Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This report makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.


### Download and unzip file used for the analysis.

```{r}

#clean repositories
rm(list=ls())

# install libraries
if (library("RCurl", logical.return = TRUE) == FALSE) {
    install.packages("RCurl")
}
if (library("lubridate", logical.return = TRUE) == FALSE) {
    install.packages("lubridate")
}
if (library("plyr", logical.return = TRUE) == FALSE) {
    install.packages("plyr")
}
if (library("lattice", logical.return = TRUE) == FALSE) {
    install.packages("lattice")
}

# assign libraries
library(RCurl)
library(lubridate)
library(plyr)
library(lattice)

# check if a data folder exists; if not then create one
if (!file.exists("data")) {dir.create("data")}

# file URL and destination file
fileUrl <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
destfile <- "./data/activity.zip"

# download the file and store the time the file was downlaod
download.file(fileUrl, destfile = destfile, mode = "wb")
dateDownloaded <- date()

# unzip the filee
activity <- read.csv(unz("./data/activity.zip", "activity.csv"))

# display top 6 rows of the file just to see that it came through 
head(activity)

```

### 1 - What is mean total number of steps taken per day?

```{r}
# calculate total number of steps pr. day using ddply and removing days with no observations
total.steps.pr.day <- na.omit(ddply(activity,~date,summarise,sum=sum(steps, na.rm = FALSE)))

#plot histogram
hist(total.steps.pr.day$sum , breaks = 10, 
          xlab = "Total number of steps per day", 
          main = "Distribution of total steps per day", 
          col = "lightblue")

# report mean
mean(total.steps.pr.day$sum, na.rm = TRUE)

# report median
median(total.steps.pr.day$sum, na.rm = TRUE)

```

### 2 - What is the average daily activity pattern?

```{r}

# calculate average number of steps pr. interval removing missing values
average.steps.pr.interval <- ddply(activity,.(interval),summarise,mean.steps=mean(steps, na.rm = TRUE))

# plot results
plot(average.steps.pr.interval$interval, average.steps.pr.interval$mean.steps, type = "l",
     xlab="Interval", ylab="Average number of steps taken")

# report max 5. minute interval
average.steps.pr.interval[average.steps.pr.interval$mean==max(average.steps.pr.interval$mean),1]

```


### 3 - Imputing missing values

```{r}
# report number of missing values
sum(is.na(activity))

#Devise a strategy for filling in all of the missing values in the dataset. 
#The strategy does not need to be #sophisticated. For example, you could use the mean/median for 
#that day, or the mean for that 5-minute interval, etc.

# find missing steps
missing.steps <- activity[is.na(activity$steps)==TRUE,]

# replace missing values by average pr. interval

# join activity and average steps pr. intervals together 
merged.dataset <- merge(activity, average.steps.pr.interval, by="interval", all.x=TRUE, suffixes=c(".Activity", ".Average"))

# replace missing values by average step pr. interval 
merged.dataset$steps <- ifelse(is.na(merged.dataset$steps), merged.dataset$mean.steps, merged.dataset$steps)

#keep only modified table 
activity.nomissing <- merged.dataset[,seq(1:3)]

#calculate number of steps each day with modified dataset
total.steps.pr.day.modified <- ddply(activity.nomissing,~date,summarise,sum=sum(steps, na.rm = FALSE))

#plot histogram
hist(total.steps.pr.day.modified$sum , breaks = 10, 
          xlab = "Total number of steps per day", 
          main = "Distribution of total steps per day (imputed values)", 
          col = "lightblue")

# report mean
mean(total.steps.pr.day.modified$sum, na.rm = TRUE)

# report median
median(total.steps.pr.day.modified$sum, na.rm = TRUE)

```

When replacing the NAs with the average of the respective interval the median and mean have the same value, indicating a more even distribution. 


### 4 - Are there differences in activity patterns between weekdays and weekends?


```{r}

# create factor weekend wekday
activity.nomissing$dayType <- as.factor(ifelse(wday(as.Date(activity.nomissing$date)) %in% c(1,7),"weekend","weekday"))

# average the steps pr.interval and pr. daytype
average.steps.pr.interval.pr.dayType <- ddply(activity.nomissing,.(interval,dayType),summarise,mean.steps=mean(steps, na.rm = TRUE))

# create a xyplot
xyplot(average.steps.pr.interval.pr.dayType$mean.steps ~ average.steps.pr.interval.pr.dayType$interval | average.steps.pr.interval.pr.dayType$dayType, layout = c(2, 1), type = "l", 
       xlab = "Time interval", ylab = "Number of steps", main = "Time series of number of steps vs time interval" )

```

Average steps taken on the weekend are much more distributed over the whole day, while during the weekdays the average spikes at interval 835
library(knitr)
library(markdown)
knit2html( )